---
title: "01. Sample preprocessing"
date: "`r format(Sys.time(),  '%d %B %Y')`"
author: "Mohamed Hassan"
output:
  html_notebook:
    theme: cerulean
    toc: true
    toc_depth: 3
editor_options:
  markdown:
    wrap: 72
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = T, results = "hide")
require("knitr")
## Adjust your diectory for the rest of the analysis

# opts_knit$set(root.dir = "/")
```

Setting a random seed to ensure results reproducibility
```{r random-seed-setting}
set.seed(12345)
```

```{r loading-libraries, results='hide'}
pacman::p_load(tidyverse, Seurat, styler, remotes, DoubletFinder, stringr, patchwork)
```

# Adding useful functions if needed to be used
```{r}
"%out%" <- Negate("%in%")
```


# 1. Reading objects
```{r}
path_to_raw_files <- "/path/to/Raw_Files/"
## The structre of the Raw_Files folder should contain subfolders, each named according to the sample identifier
## Within each sample, three files are required "barcodes, features, and matrix"

files <- list.files(path_to_raw_files)


# Iterate over each file to read the data and create Seurat objects
Mouse_sorted_ECs_list <- lapply(files, function(file) {
  # Construct the full path to the file
  file_path <- file.path(path_to_raw_files, file)
  
  # Read the data
  data <- Read10X(data.dir = file_path)
  
  # Create the Seurat object
  seurat_obj <- CreateSeuratObject(counts = data, project = file, )
  seurat_obj <- RenameCells(seurat_obj, str_c(seurat_obj$orig.ident, 
                                              rownames(seurat_obj@meta.data),
                                              sep = "_"))
  
  return(seurat_obj)
})

# Name each element of the list with the corresponding file name
names(Mouse_sorted_ECs_list) <- files

```

## Adding Mt% and Rb%

```{r}
## Adapt this function based on the scCustomize version you have (i.e in the new version its called Add_Mito_Ribo)
Mouse_sorted_ECs_list <- lapply(Mouse_sorted_ECs_list, scCustomize::Add_Mito_Ribo_Seurat, 
                          mito_name = "Mt", ribo_name = "Rb", 
                          mito_ribo_name = "Mt_Rb", species = "Human")
## Adjust the species argument and use the organism in use (e.g Mouse) 
```

## Violin plots for quality check

```{r,fig.height=10,fig.width=10}
lapply(Mouse_sorted_ECs_list, VlnPlot, 
       features = c("nCount_RNA", "nFeature_RNA", "Mt", "Rb"),
       ncol = 4)
```

## Scatter plots for quality check

```{r}
lapply(Mouse_sorted_ECs_list, FeatureScatter, 
       feature1 = "nCount_RNA", 
       feature2 = "nFeature_RNA")

```

```{r unlisting-samples_2}
for (i in 1:length(Mouse_sorted_ECs_list)) {
  object_name <- names(Mouse_sorted_ECs_list[i])
  assign(object_name, Mouse_sorted_ECs_list[[i]])
}
```


```{r}
lower_limit <- 200

## set your upper limit for each sample 
upper_limit <- 5000
```

```{r}
## For each sample

## sample1
sample1 <- subset(sample1, nCount_RNA > lower_limit & nCount_RNA <   & 
                  nFeature_RNA > lower_limit & nFeature_RNA <  & Mt < ) 
## sample2
sample2 <- subset(sample2, nCount_RNA > lower_limit & nCount_RNA <   & 
                  nFeature_RNA > lower_limit & nFeature_RNA <  & Mt < ) 
```

```{r}
## Add more samples if you have more than two, and with their corresponding names
sample_list = list("sample1"=sample1, 
                            "sample2"=sample2)
```


## Removing **ribosomal** genes
### If you want to remove specific genes before the downstream analysis (Ribosomal/Mitochondrial/..etc)

```{r}
sample_list_NoRibo <- lapply(sample_list, function(x) {subset(x, 
       features = grep(pattern = "^RP[L|S]", x = rownames(x),
                                invert = T))
})
```

## Removing **mitochondrial** genes

```{r}
raw_sample_list_NoRbMt <- lapply(raw_sample_list_NoRibo, function(x) {subset(x, 
       features = grep(pattern = "^MT-", x = rownames(x),
                                invert = T))
})
```

## Removing **MALAT1,NEAT1** genes

```{r}
raw_sample_list_NoRbMt_lnc <- lapply(raw_sample_list_NoRbMt, function(x)
  {subset(x, 
       features = grep(pattern = "^MALAT1|NEAT1", x = rownames(x),
                                invert = T))})
```

## Removing **Hemoglobin** genes

```{r}
## Hemoglobin genes are removed as they resemble contamination, depending on which species being used

# hemoglobin_genes <- c("HbA", "Hba-a2", "Hbb-b1", "Hbb-b2", "Hbb-y",
#                       "Hbb-e1", "Hbd", "Hbg1", "Hbg2")

# hemoglobin_genes <- c("HBB", "HBG2", "HBZ", "HBA2", "HBA1",
#                       "HBM", "HBD", "HBE1", "HBQ1", "HBG1")

raw_sample_list_NoRbMt_genes <- lapply(raw_sample_list_NoRbMt_lnc, function(x){
  subset(x, 
         features = rownames(x)[!(rownames(x) %in% hemoglobin_genes)])
})
```


```{r unlisting-samples}
for (i in 1:length(raw_sample_list_NoRbMt_genes)) {
  object_name <- names(raw_sample_list_NoRbMt_genes[i])
  assign(object_name, raw_sample_list_NoRbMt_genes[[i]])
}
```
###################################################################################################################################################
## Filtering by linear model (UMIs vs Genes)

#### sample1

```{r}
lin_mod <- lm(log(sample1@meta.data$nFeature_RNA) ~ log(sample1@meta.data$nCount_RNA))
sub <- lin_mod$model[lin_mod$residuals < -0.5,]
plot (log(sample1@meta.data$nCount_RNA),log(sample1@meta.data$nFeature_RNA), xlab ="log (nCount_RNA)",ylab="log (nFeature_RNA)",pch=20,col="grey")
abline (lin_mod,col="red",lwd=3)
points (sub[,2], sub[,1],pch=20,col="blue")
tokeep <- as.numeric(names(lin_mod$residuals[lin_mod$residuals >= -0.5]))
sample1 <- subset(sample1, cells = tokeep)
```

#### sample2

```{r}
lin_mod <- lm(log(sample2@meta.data$nFeature_RNA) ~ log(sample2@meta.data$nCount_RNA))
sub <- lin_mod$model[lin_mod$residuals < -0.5,]
plot (log(sample2@meta.data$nCount_RNA),log(sample2@meta.data$nFeature_RNA), xlab ="log (nCount_RNA)",ylab="log (nFeature_RNA)",pch=20,col="grey")
abline (lin_mod,col="red",lwd=3)
points (sub[,2], sub[,1],pch=20,col="blue")
tokeep <- as.numeric(names(lin_mod$residuals[lin_mod$residuals >= -0.5]))
sample2 <- subset(sample2, cells = tokeep)
```

# Removing doublets (DoubletFinder)

#### sample1

if using the new version of DoubletFinder, adjust the functions that end with _v3 

```{r}
sample1DoubletPerc <- 0.008*(ncol(sample1)/1000)
sample1DoubletPerc
```

```{r}
sample1 <- sample1 %>% NormalizeData(normalization.method = "LogNormalize", scale.factor = 10000) %>% FindVariableFeatures(selection.method = "vst", nfeatures = 2000) %>% ScaleData(vars.to.regress = c("nCount_RNA", "nFeature_RNA")) %>% RunPCA(npcs = 100)

ElbowPlot(sample1, ndims = 50)

### Determining the number of significant PCs
pct <- sample1@reductions$pca@stdev / sum(sample1@reductions$pca@stdev) * 100
cum <- cumsum(pct)
co1 <- which(cum > 90 & pct < 5)[1]
co2 <- sort(which((pct[1:length(pct) -1] - pct[2:length(pct)]) > 0.05), decreasing = T)[1] + 1
pcs = min(co1, co2)
sig_pcs <- 1:pcs

sweep.res.list_sample1 <- paramSweep_v3(sample1, PCs = sig_pcs, sct = FALSE)
sweep.stats_sample1 <- summarizeSweep(sweep.res.list_sample1, GT = FALSE)
bcmvn_sample1 <- find.pK(sweep.stats_sample1)

sample1 <- FindNeighbors(sample1, dims = sig_pcs, reduction = "pca", features = VariableFeatures(sample1))
sample1 <- FindClusters(sample1)


homotypic.prop_sample1 <- modelHomotypic(sample1@meta.data$seurat_clusters)   
nExp_poi_sample1 <- round(sample1DoubletPerc*nrow(sample1@meta.data))
nExp_poi.adj_sample1 <- round(nExp_poi_sample1*(1-homotypic.prop_sample1))


pK_sample1 = as.numeric(as.character(bcmvn_sample1$pK))
BCmetric_sample1 = bcmvn_sample1$BCmetric
pK_choose_sample1 = pK_sample1[which(BCmetric_sample1 %in% max(BCmetric_sample1))]

par(mar=c(5,4,4,8)+1,cex.main=1.2,font.main=2)
plot(x = pK_sample1, y = BCmetric_sample1, pch = 16,type="b",
col = "blue",lty=1)
abline(v=pK_choose_sample1,lwd=2,col='red',lty=2)
title("The BCmvn distributions")
text(pK_choose_sample1,max(BCmetric_sample1),as.character(pK_choose_sample1),pos = 4,col = "red")
```

```{r}
sample1 <- doubletFinder_v3(sample1, PCs = sig_pcs, pN = 0.25, pK = pK_choose_sample1, nExp = nExp_poi.adj_sample1, reuse.pANN = FALSE, sct = FALSE)
sample1 <- RunUMAP(sample1, dims = sig_pcs,reduction = "pca", verbose = FALSE)
```

```{r}
DF <- colnames(sample1@meta.data)[grep(pattern = "^DF", 
                                   colnames(sample1@meta.data))]
DimPlot(sample1, group.by=DF)
```

```{r}
sample1
singlet_cells <- sample1@meta.data %>% select(starts_with("DF")) %>% filter(. == "Singlet") %>% rownames()
sample1 <- subset(sample1, cells = singlet_cells)
sample1 
```

#### sample2

```{r}
sample2DoubletPerc <- 0.008*(ncol(sample2)/1000)
sample2DoubletPerc
```

```{r}
sample2 <- sample2 %>% NormalizeData(normalization.method = "LogNormalize", scale.factor = 10000) %>% FindVariableFeatures(selection.method = "vst", nfeatures = 2000) %>% ScaleData(vars.to.regress = c("nCount_RNA", "nFeature_RNA")) %>% RunPCA(npcs = 100)

ElbowPlot(sample2, ndims = 50)

### Determining the number of significant PCs
pct <- sample2@reductions$pca@stdev / sum(sample2@reductions$pca@stdev) * 100
cum <- cumsum(pct)
co1 <- which(cum > 90 & pct < 5)[1]
co2 <- sort(which((pct[1:length(pct) -1] - pct[2:length(pct)]) > 0.05), decreasing = T)[1] + 1
pcs = min(co1, co2)
sig_pcs <- 1:pcs


sweep.res.list_sample2 <- paramSweep_v3(sample2, PCs = sig_pcs, sct = FALSE)
sweep.stats_sample2 <- summarizeSweep(sweep.res.list_sample2, GT = FALSE)
bcmvn_sample2 <- find.pK(sweep.stats_sample2)

sample2 <- FindNeighbors(sample2, dims = sig_pcs, reduction = "pca", features = VariableFeatures(sample2))
sample2 <- FindClusters(sample2)


homotypic.prop_sample2 <- modelHomotypic(sample2@meta.data$seurat_clusters)   
nExp_poi_sample2 <- round(sample2DoubletPerc*nrow(sample2@meta.data))
nExp_poi.adj_sample2 <- round(nExp_poi_sample2*(1-homotypic.prop_sample2))


pK_sample2 = as.numeric(as.character(bcmvn_sample2$pK))
BCmetric_sample2 = bcmvn_sample2$BCmetric
pK_choose_sample2 = pK_sample2[which(BCmetric_sample2 %in% max(BCmetric_sample2))]

par(mar=c(5,4,4,8)+1,cex.main=1.2,font.main=2)
plot(x = pK_sample2, y = BCmetric_sample2, pch = 16,type="b",
col = "blue",lty=1)
abline(v=pK_choose_sample2,lwd=2,col='red',lty=2)
title("The BCmvn distributions")
text(pK_choose_sample2,max(BCmetric_sample2),as.character(pK_choose_sample2),pos = 4,col = "red")
```

```{r}
sample2 <- doubletFinder_v3(sample2, PCs = sig_pcs, pN = 0.25, pK = pK_choose_sample2, nExp = nExp_poi.adj_sample2, reuse.pANN = FALSE, sct = FALSE)
sample2 <- RunUMAP(sample2, dims = sig_pcs,reduction = "pca", verbose = FALSE)
```

```{r}
DF <- colnames(sample2@meta.data)[grep(pattern = "^DF", 
                                   colnames(sample2@meta.data))]
DimPlot(sample2, group.by=DF)
```

```{r}
sample2
singlet_cells <- sample2@meta.data %>% select(starts_with("DF")) %>% filter(. == "Singlet") %>% rownames()
sample2 <- subset(sample2, cells = singlet_cells)
sample2 
```

###################################################################################################################################################

# Merging all samples

```{r}
Merged_raw_object <- merge(sample1, c(sample2))
Merged_raw_object
```

# Adding metadata information
```{r}
df_Merged_raw_object <- Merged_raw_object@meta.data %>% as.data.frame()
## Removing unwanted columns
df_Merged_raw_object <- df_Merged_raw_object %>% select(orig.ident:Mt_Rb)

## Adding Study info

```

```{r}
Merged_raw_object@meta.data <- df_Merged_raw_object
```

```{r}
saveRDS(Merged_raw_object, "Merged_raw_object.RDS")
```






